Combining Bottom-up and top-down approaches through graph learning over interaction networks for drug-target-interaction prediction

### 2021-01-09 paper/main.tex
Update paper draft and TODO list for upcoming chapters

\section{Approach}
\subsection{Choosing appropriate train-test splits}
As shown, recent work lacks the ability to combine both top-down and bottom-up approaches for their predictions, however performing quite well on their datasets. Drug-target interaction prediction is the task to accurately predict whether for a given drug and a given protein there is a biological interaction within humans. However, when building the train-test split over compound-protein pairs for building the actual model, there are the following three options:
\begin{enumerate}
	\item Build split over drugs
	\item Build split over drug-target pairs
	\item Build split over proteins
\end{enumerate} 
In general, recent works do perform their split over the drugs or drug-target pairs (\cite{Survey2018}, CITATION). The first is more relevant for novel drugs, as it is much more likely to test a new compound than a innovative protein. However, it lies in the very nature of the used datasets, making the prediction for new drugs much easier. Thus, drugs are often built by minor variations of existing drugs, thus leading to no deviations in the functional group of that very compound (CITATION/EXAMPLE). When distributed over both train and test split, the models do not perform inductive inference and generalize, but rather implement transductive inference by just predicting the recently seen structures. Hence, when entirely new molecules are seen, the models perform much worse. \\
The same applies to splits of drug-target pairs, as all drugs were already seen, and novelty cannot be coped with.\\
As mentioned in the introduction it is quite difficult to learn suitable features from proteins. In general, attempts search for motifs in the protein sequences under usage of convolutional neural networks and filters, which is more suitable for tasks like protein function prediction, than for for drug-target interaction prediction, and lack a more in-depth hypothesis on the protein side, while investing in refined drug features. \\
Thus, building splitting over proteins is the most challenging of the three options. \\

@@

\section{Methods}
\subsection{Data}
The data for the different parts of this model were obtained from different sources. Starting with the protein-protein interactions, we fetched 11.574 proteins with over 170.000 links from STRING (\cite{STRINGv10}). As STRING provides statistical scores for each association, we filtered them as advised by a threshold of 700. For the drug-drug interaction we retrieved the drugs from \cite{Boyce2015}. Side effects were obtained from MedDRA database (\cite{MedDRA}) that hierarchically orders side effects. Links between drugs and side effects were fetched from \textit{SIDER} (\cite{SIDER}) side effects database. The intersection between these sources yielded 641 common drugs for usage in our data. For the drug-target interactions themselves, we fetched 
137.000 links from STITCH database (\cite{STITCHv5}). We provide links and methods to and for the necessary data in the provided Github repository.\\

>>>>

\section{Methods}
\subsection{Datasets}
The data for the different parts of this model were obtained from various sources. Starting with the protein-protein interactions, we fetched 11.574 proteins with over 170.000 links from STRING (\cite{STRINGv10}). For the drug-target interactions themselves, we fetched 
137.000 links from STITCH database (\cite{STITCHv5}). As both STRING and STITCH provide probability scores for each association, we filtered them as advised by a threshold of 700, thus only obtaining likely interactions.\\
For the ontology segment we utilized PhenomeNET \citep{PhenomeNET2011}, a collection of various ontologies such as Human Phenotype Ontology \citep{HPO2018}, Gene Ontology (\citet{GOoriginal2000} \citet{GOrecent2020}), Mammalian Phenotype Ontology \citep{MP2009} and numerous others. \todo{which ontologies to cite}
Side effects and their links to drugs were obtained Side Effect Resource (SIDER)\citep{SIDER} and structured according MedDRA database \citep{MedDRA}. They were mapped to PhenomeNET with aid of \textit{Phenomebrowser.net}, which provides a SPARQL query endpoint for the mentioned resources. \\

We additionally obtained molecular structure based features for drugs from \textit{SmilesTransformer}\citep{SmilesTransformer} and proteins from \textit{DeepGOPlus}\citep{DeepGoPlus}.\\

Eventually the intersection between these resources yielded \todo{num drugs} drugs and \todo{num proteins} human proteins for the training phase. We provide links to and methods for the necessary data in the provided Github repository.\\

\subsection{Problem Description}
The issue of predicting drug-target interactions can be described quite briefly: For a given drug and a given protein we want to forecast whether those interact or not. We hereby do not differentiate between activation and inhibition, and do not erect statements on the strength of the bond.\\

\subsection{Model} 
In order to build a method that incorporates both top-down and bottom-up features, we first created a model for each. In the top-down section, we used \textit{DL2vec}\citep{DL2vec2020} to obtain ontology based representations. Hereby, DL2vec constructs a graph by introducing vertices and edges for each ontology class and axiom, respectively, followed by random walks starting from each entity. These walks are eventually learned on using a Word2vec \citep{Word2vec2013} model. Thus, we pick up rich, neighbourhood focused representations for each entity, which has shown great results for representing protein function and phenotypes. \\

As we want to learn from the similarity of drug side effects and protein phenotypes we opted for a deep siamese network approach, hence learning a high-dimensional embedding emphasizing this identity by forcing a maximal cosine similarity between these embeddings. On the other hand we build a deep neural network for the molecular structure based features. \\

\subsubsection{Graph convolutional layers}
These molecular and ontology based sub-models were added to a larger graph convolutional model.
As described before, graph convolution has shown significant performance increase in a variety of tasks. While there are various methods out there we will only introduce the mose basic one here. A graph convolutional layer w.r.t. \citet{GCNConv} hereby consists of a learnable weight matrix followed by an aggregation step, formalized by

\begin{equation}
	\mathbf{X}^{\prime} = \mathbf{\hat{D}}^{-1/2} \mathbf{\hat{A}}
	\mathbf{\hat{D}}^{-1/2} \mathbf{X} \mathbf{\Theta}
\end{equation}

where for a given graph $G=(V,E)$, $\hat{A} = A + I$ denotes the adjacency matrix with added self-loops for each vertex, $D$ is described by $\hat{D}_{ii} = \sum_{j=0} \hat{A}_{ij}$, a diagonal matrix displaying the degree of each node, and $\Theta$ denotes the learnable weight matrix. Added self-loops enforce that each node representation is directly dependent on its own preceding one. Notably, the number of graph convolutional layers stacked equals the radius of relevant nodes for each vertex within the graph.\\

The update rule for each individual node is denoted by

\begin{equation}
	\mathbf{x}^{\prime}_i = \mathbf{\Theta} \sum^{N}_{j}
	\frac{1}{\sqrt{\hat{d}_j \hat{d}_i}} \mathbf{x}_j
\end{equation}

where both $\hat{d}_i, \hat{d}_j$ are dependent on the edge weights $e_{ij}$ of the graph. With simple, single valued edge weights such as $e_{ij}=1 \text{ }\forall (i,j)\in E$, all $\hat{d}_i$ reduce to $d_i$, i.e. the degree of each vertex $i$. \\

While in this initial formulation the node-wise update step is defined by the sum over all neighbouring node representations, we are able to alter this formulation for a more sophisticated message passing scheme. As described we are able to rearrange the order of activation function, aggregation and linear neural layer with this formulation as proposed by \citet{GENConv2020}:

\begin{equation}
	\mathbf{x}_i^{\prime} = \mathrm{MLP} \left( \mathbf{x}_i +
	\mathrm{AGG} \left( \left\{
	\mathrm{ReLU} \left( \mathbf{x}_j + \mathbf{e_{ji}} \right) +\epsilon
	: j \in \mathcal{N}(i) \right\} \right)
	\right)
\end{equation}

While the reordering is merely import for numerical stability, the authors claim that this alteration of the original formulation eases the vanishing gradient problem for deeper convolutional networks. \\

\subsection{Choosing a train-test splitting scheme}
In general, drug-target interaction prediction is the task of accurately predicting, whether for a given drug and a given protein there is a biological interaction within the target organism. Hereby, different training and prediction schemes lead to divergent expressiveness of the resulting model. However, when building the train-test split over compound-protein pairs for building the actual model, there are the following three options:

\begin{enumerate}
	\item Build split over drugs
	\item Build split over drug-target pairs
	\item Build split over proteins
\end{enumerate} 
In general, recent works do perform their split over the drugs or drug-target pairs (\cite{Survey2018}, CITATION). As there are hopefully many more drugs to discover, the drug split scheme both emphasizes the drug repurposing idea, by applying unseen compounds to existing targets, but also benefits from more complicated drug representations, leading to tremendous results. This performance gain is based on minor variations among large groups of pharmaceuticals, that are easy to acquire. The second scheme has knowledge on all drugs and all proteins, and is thus prone to overfitting and the same development bias. Eventually, as there only limited drug-targets \citep{Overington2006}, predicting per protein is rather counter-intuitive. As it is hard to generalize over proteins representations, we aim at reaching similar performances for both drug and protein splitting schemes. \\

In general, recent works do perform their split over the drugs or drug-target pairs (\cite{Survey2018}, CITATION). The first is more relevant for novel drugs, as it is much more likely to test a new compound than a innovative protein. However, it lies in the very nature of the used datasets, making the prediction for new drugs much easier. Thus, drugs are often built by minor variations of existing drugs, thus leading to no deviations in the functional group of that very compound (CITATION/EXAMPLE). When distributed over both train and test split, the models do not perform inductive inference and generalize, but rather implement transductive inference by just predicting the recently seen structures. Hence, when entirely new molecules are seen, the models perform much worse. \\
The same applies to splits of drug-target pairs, as all drugs were already seen, and novelty cannot be coped with.\\
As mentioned in the introduction it is quite difficult to learn suitable features from proteins. In general, attempts search for motifs in the protein sequences under usage of convolutional neural networks and filters, which is more suitable for tasks like protein function prediction, than for for drug-target interaction prediction, and lack a more in-depth hypothesis on the protein side, while investing in refined drug features. \\
Thus, building splitting over proteins is the most challenging of the three options. \\

@@

\section{Methods}
